---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```


## Overview

The **l1rotation** package implements the l1-rotation rotation criterion for factor models from Freyaldenhoven (2025).

## Installation

```{r, eval = FALSE}
# Install from CRAN
install.packages("l1rotation")

# Install latest version from GitHub
install.packages("devtools")
devtools::install_github("SimonFreyaldenhoven/l1rotation")
```

## Usage

Find a minimal example below. 
For more examples see the package [documentation](https://cran.r-project.org/package=l1rotation/l1rotation.pdf) and [vignette](https://cran.r-project.org/package=l1rotation/vignettes/documentation.html).

```{r example}

library(l1rotation)
set.seed(916) 

# Minimal example with 4 factors, where X is a 500 by 300 matrix
lf <- local_factors(X = example_data, r = 4)

# Rerun with parallel processing
lf_lambdas <- find_local_factors(X = example_data, r = 4, parallel = TRUE, n_cores = 10)

# Visualize Principal Component estimate of the loadings
lf$pc_plot

# Visualize l1-rotation loadings
lf$pc_rotated_plot
```

## Citation

Simon Freyaldenhoven."Identification Through Sparsity in Factor Models: the l1-rotation criterion" [Working Paper](https://simonfreyaldenhoven.github.io/papers/factor_rotation.pdf),
February 2025.

Simon Freyaldenhoven, Ryan Kobler. "`l1rotation` package." Code and data repository at https://github.com/SimonFreyaldenhoven/l1rotation, March 2025.
